\chapter{Introduction}\label{sec:introduction} % approx. 2 pages


\section{Motivation}\label{sec:introduction-motivation}
Shared micromobility services, such as electric scooter sharing, are a large, growing part of urban and suburban mobility \cite{oeschger_micromobility_2020}.
Shared micromobility offers an ecologically friendly and convenient way to bridge short distances \cite{liao_electric_2022}.
They address the common problem of the \textit{first and last mile gap}, which describes the distance between public transport stations and the trip's origin and destination \cite{liao_electric_2022}.
The growing popularity of micromobility sharing services comes from their flexibility, ease of use, low environmental impacts, and the possibility of reducing private car rides \cite{jaber_preferences_2023}.
However, this popularity brings significant challenges for service providers aiming to maintain high service quality.
The spatiotemporal dynamics of urban mobility, particularly the flow patterns of commuters, systematically create significant fleet imbalances within micromobility systems \cite{wu_challenges_2020}.
These imbalances lead to a poor user experience, as evident in the unavailability and overcrowding of vehicles \cite{wu_challenges_2020}.
To counteract these effects, operators employ rebalancing strategies \cite{dellamico_bike_2014}.
Rebalancing is a common approach to maintaining a balanced fleet of vehicles and, therefore, providing a high service quality.
Rebalancing refers to the process of relocating vehicles from low-demand areas to high-demand areas to increase satisfied user demand.
This can be achieved through two primary modes: operator-based rebalancing, where service crews manually move vehicles using trucks, and user-based rebalancing, where users are incentivized to start or end their trips at specific, over- or under-supplied locations \cite{emami_integrated_2024}.
Effectively managing these strategies is critical to maximizing vehicle availability, user satisfaction, and keeping operational costs low.

\section{Research Gap and Objectives}\label{sec:introduction-reserach_gap}
Various rebalancing methods exist, but current approaches suffer from significant limitations.
Traditional strategies, particularly those based on mathematical optimization, struggle to scale for large real-world systems due to the exponential growth in computational complexity inherent in solving routing and allocation problems.
Moreover, heuristic and optimization methods often depend on static or pre-computed demand models, which can limit their responsiveness to sudden changes in demand caused by public events or other real-time dynamics.
Demand can peak unexpectedly due to public events or similar, which static rebalancing strategies can not account for.
This makes such approaches slow to adapt to the highly dynamic nature of urban mobility.
Additionally, such strategies struggle to scale for large real-world systems.

\glsfirst{rl} approaches have emerged due to the significant relevance of artificial intelligence.
They offer more dynamic strategies than traditional approaches.
However, a critical research gap remains: the vast majority of existing studies address operator-based and user-based rebalancing in isolation, without integrating them into a unified framework.
This research gap is significant because the rebalancing problem naturally follows a hierarchy.
It involves both strategic decisions regarding fleet distribution between city districts and tactical decisions for specific regions.
This structure supports the implementation of Hierarchical Reinforcement Learning (HRL), which is explicitly designed to decompose problems into interconnected sub-policies.
This clear separation between the two rebalancing approaches overlooks the potential of a combined strategy that can benefit from the advantages of both approaches while mitigating their limitations when properly combined.
Furthermore, many \gls{rl} rebalancing strategies focus on either city-wide or local neighborhood optimization, but both rely on each other, which, combined, could significantly enhance service quality.\\

This thesis addresses these identified research gaps by proposing an integrated system that combines both rebalancing modes.
The primary objectives of this work are:
\begin{goal}\label{goal:design}
    Design a \glsfirst{hrl} framework that integrates operator- and user-based rebalancing by decomposing the problem into spatially and functionally distinct sub-problems.
\end{goal}
\begin{goal}\label{goal:implementation}
    Implement the designed \gls{hrl} framework by developing agents with specific \gls{rl} algorithms and creating a custom simulation environment driven by real-world demand data.
\end{goal}
\begin{goal}\label{goal:evaluation}
    Evaluate the framework's effectiveness by systematically benchmarking its performance against multiple baselines, using metrics for service quality, operational cost, and fleet distribution.
\end{goal}



\section{Structure}\label{sec:introduction-structure}
The remainder of this thesis is structured as follows:
\autoref{sec:background} introduces the key foundations for fleet optimization.
\autoref{sec:related-work} surveys related work on optimizing shared micromobility services through rebalancing algorithms.
\autoref{sec:design} describes the available data for this study, defines the rebalancing problem for a combined user- and operator-based rebalancing system, and details the design choices of the proposed \gls{hrl} framework.
\autoref{sec:implementation} focuses on the implementational details of the core components of the proposed rebalancing framework.
\autoref{sec:evaluation} details the hyperparameter optimization conducted for the \gls{rl} agents and presents and discusses the results of evaluating the \gls{hrl} framework against multiple baselines.
Lastly, \autoref{sec:conclusion} summarizes the key findings of this thesis and provides recommendations for future work in this research area.