\chapter{Related Work}\label{sec:related-work} % approx. 13 pages
The literature review presented in this chapter follows a systematic methodology to identify, organize, and analyze relevant research regarding rebalancing strategies for shared micromobility systems.
The process was initiated with a comprehensive search on Google Scholar, utilizing keywords such as \textit{micromobility}, \textit{fleet optimization}, \textit{rebalancing}, \textit{e-scooter sharing}, and \textit{bike sharing}.
This initial query was subsequently expanded through citation chaining to uncover additional publications, yielding approximately 70 potentially relevant research papers.

To manage this large volume of literature, a preliminary screening was performed.
The Large Language Model (LLM) ChatGPT (\textit{o3-mini-high}) was utilized to generate standardized summaries of each paper from its PDF file.
This approach facilitated an efficient initial assessment of each paper's relevance, which reduced the selection to approximately 25 core research papers for in-depth review.

The selected papers were systematically organized and categorized using Zotero.
This process led to a structured classification system based on key attributes, including the type of sharing system (e-scooter sharing (ESS) or bike sharing (BSS)) and the employed rebalancing mode.
These modes are defined as operator-based (OB), user-based (UB), or autonomous (AU).

This review identified three primary categories of rebalancing algorithms: Reinforcement Learning, mathematical optimization, and heuristic-based approaches.
The following sections explore the most significant contributions within each of these categories, supplemented by tables that summarize key papers published between 2018 and 2025.
Furthermore, a dedicated section examines the simulation environments implemented to evaluate rebalancing strategies.
The chapter concludes with a summary of the findings and a critical analysis of remaining gaps in the literature, thereby highlighting the contribution of this thesis.
\\

\section{Mathematical Optimization Approaches}\label{sec:related-work_mathematical-optimization}

Lee et al. \cite{lee_battery_2024} investigate the operational management of free-floating shared e-scooter systems, addressing the challenges of battery swapping, rebalancing, and staff routing. 
The authors formulate a Mixed-Integer Linear Programming (MILP) problem comprising decision variables such as the selection of scooters for service, battery swap operations, rebalancing staff routing, and the quantity of e-scooters to be relocated between zones.
This MILP problem aims to maximize total profit, which is derived from the revenue generated by satisfied user demand and the costs associated with rebalancing and battery swapping operations.
The MILP formulation incorporates several constraints, including routing viability, staff capacity, and vehicle limitations. 

Given the complexity of this large-scale MILP problem, a Clustered Iterative Construction Approach (CICA) is proposed, which divides the solution process into two phases.
The initial phase of CICA involves clustering multiple service area regions, each assigned to a single staff member, thereby decomposing the overall problem complexity.
A minimal spanning tree (MST) approximates routing costs between areas.
In contrast, operational costs within each region are estimated based on the number of battery swaps and rebalancing operations, taking into account the state of charge and regional demand.
Imbalances between the available scooters and demand are also incorporated through a penalty term.
The second phase comprises three distinct steps. 
Initially, the algorithm generates an optimal sequence of regions to visit for each cluster.
Subsequently, the algorithm assigns specific operations for each of the stops within the previously generated sequences, such as pick-up, drop-off, and battery swap. 
This is done iteratively until the constraints are exceeded or the estimated demand is fully met. 
In the final step, the sequences with the highest profit are chosen. 
The results demonstrate that the proposed CICA algorithm consistently delivers near-optimal solutions, outperforming baselines in both effectiveness and efficiency, due to its reduced computational time.
\\

Saum et al. \cite{saum_optimizing_2024} propose a framework for optimizing dockless shared e-scooter rebalancing operations that utilizes machine learning techniques in combination with mathematical optimization. 
Their rebalancing approach builds upon a Gradient Boosting demand prediction model. 
This prediction model forecasts the net trip gap between pick-ups and drop-offs. 
The parameters of this Gradient Boosting regressor are optimized using Bayesian optimization. 
To generate various demand scenarios, a Monte Carlo simulation is employed in conjunction with a Seasonal Generalized Autoregressive Conditional Heteroskedasticity (SGARCH) framework. 

The rebalancing problem is formulated as an Integer Linear Programming (ILP) model. 
This ILP model aims to identify optimal routes for service workers to allocate e-scooters.
A generalized cost function is introduced based on the driving distance, pick-up and drop-off costs, and penalties for unmet demand and overcrowding. 
The algorithm accounts for various e-scooter states, including low battery levels to trigger recharging, faulty e-scooters that require replacement, and usable e-scooters available for rebalancing. 
For small-scale problems, the ILP is solved using the GNU Linear Programming Kit (GLPK).
In contrast, more complex problems are addressed using a hybrid Ant Colony Optimization–ILP (ACO–ILP) algorithm. 
This hybrid approach comprises two stages. 
The ACO generates promising routes during the initial stage based on pheromone trails and visibility factors. 
In the following stage, the ILP solver optimizes the pick-up and drop-off operations.
In summary, this work proposes a more complex demand forecasting model, a Monte Carlo simulation, and an SGARCH model. 
However, this approach heavily depends on historical data, which does not account for demand dynamics and fluctuations and only provides a static planning horizon. 
Consequently, this approach exhibits limited responsiveness to unpredictable demand.
\\ 

Emami and Ramezani \cite{emami_integrated_2024} provide an integrated user- and operator-based rebalancing framework for an e-bike sharing service operating in Manhattan, USA. 
Operator-based rebalancing involves relocating e-bikes by a truck crew from low-demand to high-demand areas, alongside on-the-go battery-swap operations for recharging.
User-based rebalancing, facilitated by monetary incentives, exclusively provides recommendations for dropping off e-bikes in low-demand areas, without including pick-up recommendations.
The probability of users choosing to drop-off at a location different from their anticipated destination is derived using a willingness function.
This willingness function quantifies the gain a user receives from dropping off at a specific location rather than adhering to their anticipated route.
This function incorporates factors such as the incentive level, additional walking time, and designated drop-off locations.
Utilizing a logit model, this willingness function calculates the probability that a user will accept the offered incentive.
Emami and Ramezani modeled the problem as a Mixed Binary Rebalancing Problem (MBRP).
The objective of this model is to minimize the cost function, which comprises truck travel costs, battery swapping costs, user incentive expenses, and a penalty for unmet demand.
This MBRP is formulated as a Mixed Binary Linear Program (MBLP).
The authors propose a branch-and-bound algorithm to solve the MBLP, yielding a near-optimal solution.
The primary advantage of this integrated approach is that it combines user-based and operator-based rebalancing within a single framework, whereas existing literature only covers these approaches in isolation.
In their simulation, the authors utilize real taxi demand data from Manhattan, applying random adjustments to predicted locations and demand volume fluctuations to account for uncertainty and real-world variability.
However, the reliance on taxi demand data may impact the effectiveness of the proposed rebalancing system, as taxi demand can differ significantly from that for shared e-bikes.
The proposed rebalancing framework is compared to three baselines: no charging and no rebalancing, Operator-based charging and no rebalancing, and Operator-based charging and rebalancing.
A comprehensive comparison with other rebalancing algorithms from prior research is absent.
\\

Rebalancing algorithms employing mathematical optimization techniques primarily concentrate on operator-based rebalancing, particularly addressing the routing problem of service vehicles.
These approaches frequently encounter challenges with larger sharing systems, leading much of the research to employ simplified representations of the sharing system or to focus solely on a subset of the operational area.
Problem formulation is commonly achieved using MILP or LP, subsequently solved by algorithms such as Branch and Bound (BB), Interior Point Method (IPM), Coin-or Branch and Cut (CBC), or similar techniques.
Some research combines traditional solving algorithms with heuristics, thereby enabling the enhancement of problem definitions for more complex models.
However, these methods are often constrained by their reliance on static demand forecasts, simplified system representations, and fixed rebalancing intervals.
These limitations consequently restrict the scalability and responsiveness of such systems to real-time demand fluctuations.
Such frameworks should prioritize capturing demand more dynamically and decomposing the rebalancing problem to enhance scalability.\\

\begin{table}[!htb]
    \centering
    \begin{tabularx}{.8\textwidth}{cXXYY}
        \toprule
        & \textbf{Model} & \textbf{Solver} & \makecell{\textbf{Sharing}\\\textbf{System}} & \makecell{\textbf{Rebalancing}\\\textbf{Mode}}\\
        \midrule
        \cite{villa-zapata_incorporating_2024} & MILP & CBC & ESS, BSS & OB\\
        \cite{lee_battery_2024} & MILP & CICA & ESS & OB\\
        \cite{saum_optimizing_2024} & MILP & ACO–ILP & ESS & OB\\
        \cite{emami_integrated_2024} & MBRP & BB & BSS/ESS & UB \& OB\\
        \cite{jimenez-merono_optimization_2024} & MILP &  BB & BSS & OB\\
        \cite{paparella_optimization-based_2023} & MILP & - & ESS/BSS & OB\\
        \cite{sanchez_simulation_2022} & LP & IPM & BSS & AU\\
        \cite{schroer_online_2019} & MILP & CBC & CSS & OB\\
        \cite{shuai_wang_towards_2019} & LP & - & BSS & OB\\
        \cite{s_k_ghosh_dynamic_2017} & MILP & LDD & BSS & OB\\
        \bottomrule
    \end{tabularx}
    \caption{Summary of Mathematical Optimization Approaches in Rebalancing Literature. This table categorizes key research papers from 2018 to 2025 that utilize mathematical optimization for rebalancing shared mobility systems. It details the specific optimization model (e.g., MILP, LP), the solver used, the type of sharing system (e-scooter or bike sharing), and the rebalancing mode. The summary highlights that existing literature in this domain predominantly focuses on operator-based (OB) rebalancing and often employs Mixed-Integer Linear Programming (MILP) formulations.}
    \label{tab:related-works_mathematical optimization}
\end{table}

\section{Heuristic-based Approaches}\label{sec:related-work_heuristics}
Shen et al. \cite{shen_enhancing_2024} introduce a framework for evaluating and optimizing the greenhouse gas emissions of shared e-bike systems, including both docked and dockless variants. 
Part of their work is a rebalancing algorithm based on Ant Colony Optimization (ACO) that focuses on optimizing routes for their service vehicle. 
Initially, the framework uses hierarchical clustering and a Voronoi diagram to create zones based on demand similarity. 
The rebalancing problem is decomposed into two subproblems, one for intra-zone routing and the other for inter-zone routing. 
In the first phase, which focuses on intra-zone rebalancing, the algorithm determines pick-up and drop-off locations based on the current number of available bikes and the expected demand for the next day. 
This is done for each zone, treating the geometric center of the zone as the pick-up and drop-off points as a Traveling Salesman Problem (TSP) that ACO solves. 
The second phase addresses the routing of service vehicles between these zones, where ACO is again employed to determine the optimal path.
While this two-phase algorithm provides a robust mechanism for fleet rebalancing, the authors acknowledge its significant computational intensity.
The authors suggest that the high computational time demonstrated on high-end servers may pose a significant challenge when scaling the system to larger urban areas or integrating extended datasets.
Another limitation of this rebalancing approach is its extensive reliance on static parameter settings, such as the acceptable walking distance or demand rate.
This use of static values reduces the framework's ability to adapt to the dynamic fluctuations characteristic of real-world micromobility services.
\\

Kim et al. \cite{kim_optimal_2023} developed a near-optimal rebalancing strategy for e-scooter systems utilizing a Genetic algorithm (GA) approach. 
The proposed system builds upon previous work on rebalancing bike-sharing systems and adapting them to perform similarly for e-scooter-sharing systems. 
The authors begin by formulating the rebalancing problem as an MILP that aims to distribute the available e-scooters to meet the target demand.
The secondary objective is to minimize the expected time required for rebalancing.
The target demand is defined by the forecasted demand during the morning peak, and the rebalancing is performed in a low-demand period in the early morning to prepare for the morning peak. 
Consequently, this approach does not involve dynamic rebalancing and cannot adapt to unexpected changes in demand throughout the day.
The algorithm employs standard GA operators, including roulette wheel selection, sequential recombination crossover, and mutation, to balance exploration and exploitation.
Through detailed parameter tuning, the authors demonstrate that the GA outperforms traditional solvers, such as CPLEX, by significantly reducing computational time. 
A comparative analysis reveals that the GA-based algorithm outperforms traditional solvers, such as CPLEX, by significantly reducing computational time.
The algorithm efficiently rebalances e-scooters to match demand and supply with fewer rebalancing vehicles.
However, for large-scale sharing systems, the computational time of this GA approach increases exponentially, making it infeasible for such systems to perform in real-time.
Another limitation is the use of a simplified system model, which relies on Euclidean distances and constant vehicle speeds, potentially impacting the accuracy of the results in a real-world context.
\\

Tolomei et al. \cite{tolomei_benefits_2021} propose a data-informed framework to optimize e-scooter sharing fleets. 
Their framework consists of two main components: a predictive demand model and a relocation heuristic. 
The framework was developed using open data from the US cities of Louisville and Austin.
Two demand models were evaluated: a simple baseline using historical averages and a more advanced 3D-CLoST model combining a Convolutional Neural Network (CNN) with Long Short-Term Memory (LSTM) layers.
Based on the demand predictions for the next hour, a simple greedy heuristic assigns relocation tasks to the nearest workers to address the most significant predicted imbalances between supply and demand.
The results showed that rebalancing guided by the more accurate 3D-CLoST model significantly increased the number of satisfied trips.
This finding suggests that even a simple rebalancing heuristic can be made more effective with precise demand forecasting.
However, the greedy rebalancing logic is not benchmarked against more sophisticated optimization methods. Its myopic nature may not adequately capture the complex system dynamics, limiting its applicability to real-world sharing systems.
Nevertheless, this research demonstrated the importance of advanced, effective demand forecasting as the foundation of a successful rebalancing strategy.
\\

Heuristic-based rebalancing algorithms have demonstrated their value as a computationally efficient alternative to exact mathematical optimization techniques, producing near-optimal fleet distributions in significantly reduced time.
The review of the literature reveals that such rebalancing algorithms predominantly focus on operator-based rebalancing under static demand conditions.
A common strategy is to frame the task as a vehicle routing problem to find optimal routes for staff vehicles to rebalance the sharing system, frequently by formulating it as a TSP.
A recurring theme in these approaches is that the rebalancing problem is decomposed into separate subtasks, most commonly the independent optimization of each service vehicle's route, allowing a single algorithm to detect regions of under- or oversupply, propose fleet relocations, and compute vehicle tours.
This monolithic approach, which merges imbalance detection, fleet distribution, and route planning into a single optimization process, can compromise the overall quality of the solution.
The inherent trade-offs between these different objectives often result in a loss of precision, leading to suboptimal outcomes.
While concentrating on a single one of these subproblems can yield more accurate and efficient results for that specific task, this narrow focus can overlook the broader system dynamics.
\\

\begin{table}[!htb]
    \centering
    \begin{tabularx}{.8\textwidth}{cXYY}
        \toprule
         & \textbf{Heuristic} & \makecell{\textbf{Sharing}\\\textbf{System}} & \makecell{\textbf{Rebalancing}\\\textbf{Mode}}\\
        \midrule
        \cite{shen_enhancing_2024} & ACO & BSS & OB\\
        \cite{kim_optimal_2023} & GAs & ESS & OB\\
        \cite{tolomei_benefits_2021} & greedy & ESS & OB\\
        \cite{sessa_online_2021} & greedy & BSS/ESS & OB\\
        \cite{wu_challenges_2020} & greed\&match & BSS & UB\\
        \cite{peiyu_yi_rebalancing_2019} & greedy & BSS & UB\\
        \bottomrule
    \end{tabularx}
    \caption{Summary of Heuristic-Based Approaches in Rebalancing Literature. This table provides an overview of recent studies that apply heuristic and metaheuristic algorithms to the fleet rebalancing problem. It lists the primary heuristic employed (e.g., ACO, GAs, greedy), the type of sharing system, and the rebalancing mode addressed in each study. A key takeaway from this summary is that heuristic-based methods are most commonly applied to static, operator-based (OB) vehicle routing problems.}
    \label{tab:related-works_heuristics}
\end{table}

\section{Reinforcement Learning Approaches}\label{sec:related-work_rl}
Losapio et al. \cite{losapio_smart_2022} present the Multi-Agent Deep Reinforcement Learning system called E-Scooter Balancing Deep Q-Network (ESB-DQN). 
This system consists of two agents: a pick-up and a drop-off agent.
These agents propose alternative zones for users, limited to adjacent 1-hop areas, by offering incentives to encourage deviations from planned trips.
The operational area for the dockless e-scooter service is divided into 279 zones, each measuring 200 x 200 meters. 
Both agents are implemented as Deep Q-Networks (DQNs), utilizing an $\epsilon$-greedy strategy combined with experience replay, following the Rainbow DQN architecture. 
The state representation consists of three vectors of size $1 \times n$, where $n$ is the number of zones, representing: the number of available scooters, the expected deficit or surplus from demand forecasting, and a binary flag indicating zones involved in the current trip.
Upon each trip request, the pick-up and drop-off agents recommend an alternative adjacent zone. 
The simulator then processes the trip request and the associated recommendation before transitioning to the next state.

The ESB-DQN is evaluated in a simulation environment based on the ODySSEUS simulator introduced by Ciociola et al. \cite{ciociola_e-scooter_2020}, which will be discussed in more detail in \autoref{sec:related-work_simulator}. 
The work accounts for user-based rebalancing, operator-based rebalancing, and battery swaps. 
The system uses a priority queue to handle both rebalancing and battery swapping.
It assigns each crew of workers a schedule that is modeled and optimized by formulating a TSP and solving it.
Rebalancing and battery swapping operations occur within a specified time window, typically standard working hours (e.g., 9:00 a.m. to 5:00 p.m.).
\\

Cederle et al. \cite{cederle_fairness-oriented_2025} developed an RL approach to enhance the fairness of micromobility sharing services. 
They address the problem that earlier rebalancing algorithms primarily focus on efficiency and profit maximization, rather than a fair and equitable fleet distribution.
Such rebalancing algorithms tend to prioritize rebalancing to wealthier areas while excluding poorer regions, due to the differing demand rates in those areas.
To counteract this bias, Cederle et al. introduced a fairness metric and integrated it, alongside traditional efficiency measures, into Q-learning agents that perform rebalancing tasks. 
Instead of employing a classic Multi-Agent Reinforcement Learning (MARL) approach, they utilize one agent for each zone. 
Each agent undergoes independent training, which breaks the MARL problem into multiple single-agent subproblems. 
They formulated the problem first as a Multi-Agent MDP and factored it into single-agent subproblems, which enables them to adapt it to other cities or services with significantly less effort than alternative methods. 
The fairness metric employed is the Gini index, which is combined with a weighting parameter to balance the optimization of fairness and efficiency. 
The reward function thus incorporates this metric, along with penalties for unmet demand, rebalancing costs, and vehicle overcrowding.
Each agent chooses between thirteen different actions, representing a need or an oversupply of vehicles in steps of size five, ranging from -30 to 30. 
The agents operate independently, and the system assumes access to a central vehicle depot to resolve any net vehicle deficits or surpluses across the system.
By combining fairness and efficiency metrics, they reduced inequality by over 80\% while incurring a 30\% rise in costs.
Still, the trade-off between fleet distribution equity and incurred costs can be adjusted by modifying a weighting parameter.

Cederle et al. introduce the first operational RL framework that integrates fairness metrics to balance shared micromobility fleets.
Their method provides a flexible approach to embedding fairness.
Still, the framework remains quite simple: it relies on a basic demand model, performs static rebalancing only twice a day, and applies only a single fairness metric.
This simplicity highlights the need for more realistic and dynamic frameworks that incorporate fairness measures in constantly changing environments.
\\

Pan et al. \cite{pan_deep_2019} introduced the Hierarchical Reinforcement Pricing (HRP) algorithm, a method for user-based rebalancing in large-scale micromobility sharing systems.
Recognizing the scalability challenges of traditional RL, the framework employs an architectural \textit{divide-and-conquer} decomposition within a single-agent Deep Deterministic Policy Gradient (DDPG) framework, rather than a conventional multi-level agent hierarchy.
The entire rebalancing problem is modeled as one large-scale Markov Decision Process (MDP).
Its state representation consists of the current bike supply, historical demand, arrivals, the ratio of unmet to met demand, and the remaining incentive budget.
The agent's action is a continuous vector of region-specific incentive prices.

The core innovation of HRP lies in its critic network, which decomposes the total Q-value into an additive combination of regional estimators.
This design addresses two key challenges.
First, to capture temporal dependencies, each regional estimator incorporates a Gated Recurrent Unit (GRU) to process time-series patterns in demand and supply.
Second, to account for spatial dependencies, a localized bias-correction module is embedded for each region, taking the state of its direct neighbors as additional input to model the flow of users between them.
This decomposed critic architecture provides a more stable and accurate learning signal, which in turn increases sampling efficiency and stabilizes convergence during training.

Although results show significant improvements over baselines, the approach has limitations.
It relies on predefined assumptions about user willingness functions and historical data distributions, which may not capture real-world variability or unexpected demand shifts.
\\

Xie et al. \cite{xie_censored_2023} used a different approach than the previously discussed papers. 
They proposed a censored semi-bandit model for allocating bikes from a docked bike-sharing system to other zones. 
The operating area is divided into 12 different zones. 
The authors formulate a censored semi-bandit problem where each arm corresponds to one of the zones. 
The semi-bandit formulation, unlike a standard Multi-Armed Bandit (MAB), enables the model to select multiple arms (zones) in each round.
In each decision round, which is done once a day, the model allocates bikes to each zone. 
The primary objective is to maximize the cumulated number of successful bike pick-ups.
This model employs an Upper Confidence Bound (UCB)- based policy, which adds an exploration bonus to zones where demand is uncertain. 
A key contribution of this work is that the authors address the common problem of micromobility service data being censored, which means that the available historical trip data only contains data that has met demand and is missing data on unmet demand. 
This missing information is crucial for evaluating rebalancing strategies, as their primary goal is to minimize such service failures.
In this research paper, the authors use a nonparametric estimator to estimate the survival function of demand by comparing the frequency of observed pick-ups above a threshold to the number of opportunities to observe such demand. 
This estimate is then used as the exploration bonus for zones. 
They also proposed two extensions of their MAB model. The first extension incorporates contextual data, such as weather data affecting user demand, to enhance relocation. The second extension considers relocation costs to balance the benefits and costs.
The proposed UCB policy is compared to benchmark strategies, such as a greedy or naive policy that ignores censored data. 

The results demonstrate that the proposed UCB-based policy achieves a higher number of successful pick-ups compared to the baseline models.
A limitation of the model, inherent to the MAB formulation, is the assumption of a static underlying demand distribution.
This simplifies reality and restricts the model's capacity to adapt dynamically to changing demand patterns after the initial exploration phase.
\\

Schofield et al. \cite{schofield_handling_2021} develop an incentive-based rebalancing approach, evaluating three distinct policy-gradient RL algorithms: Trust Region Policy Optimization (TRPO), Proximal Policy Optimization (PPO), and Actor-Critic using Kronecker-Factored Trust Region (ACKTR).
The problem is framed as an MDP where the service area is partitioned spatially and time is discretized into hourly intervals.
The state is composed of vectors representing the number of bikes per zone, demand forecasts, remaining budget, and unsatisfied demand.
The action is a vector of incentive values for each zone, and the reward is defined by the service level (i.e., the fraction of satisfied demand).
The performance of the three algorithms is compared against each other and two baselines: a lower bound with no incentives and an unconstrained upper bound with maximum incentives.
While all three algorithms achieve similar final performance, PPO demonstrates the fastest convergence.
The authors note that the benefits of rebalancing are most pronounced in scenarios with a limited supply of vehicles.
A limitation noted is the simplicity of the simulation environment, which utilizes a discrete network and hourly time steps while omitting key real-world factors such as vehicle state-of-charge and variable trip durations.
\\

Li et al. \cite {li_dynamic_2018} propose an approach that utilizes Spatiotemporal Reinforcement Learning (STRL) to optimize the fleet distribution in bike-sharing services dynamically. 
The core methodology begins with a two-stage clustering process, called Inter-Independent Inner-Balance (IIIB), in which geographically close stations with similar demand characteristics are grouped into balanced clusters.
Each cluster is then treated independently to reduce complexity. 
The second stage incorporates predictive modeling through two components: an O-model, which uses a similarity-based k-nearest neighbor approach to forecast pick-up demand, and an I-model, which predicts drop-off demand by probabilistically tracking bike transitions across regions. 
A central element of their approach is a deep Q-learning-based framework, in which individual service vehicles, modeled as agents, utilize historical observations and real-time predictive insights to determine optimal repositioning actions.
This implicitly coordinates their actions by considering each other's states and planned moves. 
Evaluation results demonstrate the effectiveness of this approach, which significantly reduces unmet demand compared to heuristic repositioning strategies. 
However, the proposed model also reveals limitations. 
While the authors implement spatio-temporal pruning heuristics to reduce computational complexity, scalability remains constrained by the exponential growth of state-action spaces as system size increases. 
Sensitivity to prediction accuracy for demand forecasts and potential inefficiencies resulting from asynchronous decision-making among multiple agents can impact the effectiveness of the framework.
\\

%The surveyed literature discussed both single-agent and multi-agent solutions for rebalancing of shared micromobility fleets.
%Single-agent approaches either use only one agent for the entire system or, more commonly, decompose the rebalancing problem into multiple subproblems, which are solved independently by separate agents.
%Similar to Loaspio et al. \cite{losapio_smart_2022}, where one agent focused on pick-up and one agent focused on drop-off locations, or as done by Pan et al. \cite{pan_deep_2019}, which uses one agent for each geographic zone.
%This factorization of the rebalancing problem reduces computational complexity and simplifies the overall RL framework.
%However, these simplifications come at the cost of missing interdependencies between geographical zones or interacting behaviour.
%Multi-agent approaches, on the other hand, model multiple decision-makers that learn concurrently, being aware of each other and taking into account the actions of different agents.
%This collaborative approach can capture more realistic system dynamics of shared micromobility services, such as shared resources, coordinated incentives, or a shared strategy for rebalancing.
%Li et al. \cite{li_dynamic_2018} described an approach that built a MARL framework, where each agent corresponded to one of the service vehicles, which relocate the shared vehicles. 
%Their approach utilized a shared state representation, which also informed the agents about each other's intended actions, enabling them to learn and balance the whole fleet collaboratively.
%While MARL frameworks better reflect real-world scenarios, they also require more computational resources and tend to have issues with convergence stability and scalability.
%Ultimately, the decision between single-agent and multi-agent approaches for rebalancing shared micromobility fleets remains a trade-off between their advantages and disadvantages, with neither approach proving superior to the other.


\begin{table}[!htb]
    \centering
    \begin{tabularx}{.8\textwidth}{cXYY}
        \toprule
         & \textbf{Algorithm} & \makecell{\textbf{Sharing}\\\textbf{System}} & \makecell{\textbf{Rebalancing}\\\textbf{Mode}}\\
        \midrule
        \cite{cederle_fairness-oriented_2025} & Q-Learning & BSS/ESS & UB\\
        \cite{liang_dual_2024} & DQ-Network & BSS & OB\\
        \cite{yun_price_2024} & PPO & ESS & UB\\
        \cite{xie_censored_2023} & MAB & BSS & OB\\
        \cite{losapio_smart_2022} & DQ-Network & ESS & UB \& OB\\
        \cite{schofield_handling_2021} & PPO & BSS & UB\\
        \cite{pan_deep_2019} & DDPG (HRL) & BSS & UB\\
        \cite{duan_optimizing_2019} & DDPG & BSS & UB\\
        \cite{li_dynamic_2018} & DQ-Network & BSS & OB\\
        \bottomrule
    \end{tabularx}
    \caption{Summary of Reinforcement Learning Approaches in Rebalancing Literature. This table summarizes significant research employing Reinforcement Learning (RL) for shared mobility rebalancing. Each entry specifies the core RL algorithm (e.g., Q-Learning, DQN, PPO, DDPG), the type of sharing system, and the rebalancing mode. This review reveals that RL-based approaches in the literature have primarily focused on user-based (UB) incentive mechanisms, with fewer studies addressing operator-based or integrated strategies.}
    \label{tab:related-works_reinforcement-learning}
\end{table}

In summary, existing literature on RL-based rebalancing algorithms predominantly emphasizes user-based strategies, seeking optimal incentives for users to pick-up and drop-off vehicles at alternative locations.
Conversely, research explicitly targeting operator-based rebalancing remains limited, with even fewer studies investigating hybrid approaches combining both user and operator rebalancing.
The most frequently applied methodologies fall into two primary categories.
The first is value-based algorithms, such as Deep Q-Networks (DQN) and Q-learning.
The second is policy-gradient methods, including Proximal Policy Optimization (PPO), Trust Region Policy Optimization (TRPO), and Actor-Critic using Kronecker-Factored Trust Region (ACKTR).
Recent studies have explored Hierarchical Reinforcement Learning (HRL) and Multi-Agent Reinforcement Learning (MARL) frameworks to decompose complex rebalancing tasks into manageable subtasks addressed by multiple agents.
The choice between factorizing the problem into independent single-agent tasks and employing a fully multi-agent framework represents a fundamental design trade-off between computational scalability and the ability to capture complex inter-agent dynamics.
Such hierarchical and multi-agent approaches have demonstrated enhanced scalability and effectiveness in managing large-scale shared mobility services \cite{pan_deep_2019, li_dynamic_2018}.

Several studies suggest that RL-based methods can achieve superior performance over mathematical optimization and heuristic-based strategies.
However, comparative evaluations remain limited, restricted to only a few published studies \cite{li_dynamic_2018, pan_deep_2019, liang_dual_2024}.
Furthermore, fairness considerations in fleet distribution have been almost never addressed, with Cederle et al. \cite{cederle_fairness-oriented_2025} presenting the only framework that incorporates fairness metrics to improve fleet distribution.
Comprehensive analyses balancing operational costs against fairness considerations are notably absent in current research.
Additionally, existing approaches predominantly utilize historical demand data or coarse-grained forecasts, resulting in limited responsiveness to real-time demand fluctuations.
Consequently, future research should prioritize the development of dynamic demand forecasting methods, thereby enhancing the adaptability of rebalancing algorithms to shifting demand conditions.
\\

\section{Simulation Environments}\label{sec:related-work_simulator}
Developing and evaluating rebalancing frameworks is challenging due to the complex dynamics of urban demand and the high costs associated with operational trials.
Consequently, high-fidelity simulation environments can be utilized to evaluate rebalancing methods before they are deployed in the real world.
Simulation enables safe and repeatable testing of rebalancing frameworks under various scenarios.
This section introduces the most commonly used simulation frameworks for evaluating rebalancing frameworks.
\\

\subsubsection{SUMO}
The Simulation of Urban Mobility (SUMO) is an open-source, microscopic traffic simulation platform widely used in mobility research \cite{alvarez_lopez_simulation_2025}.
SUMO models individual vehicle movements with high granularity, employing car-following and lane-changing models to reproduce realistic traffic dynamics.
Due to its modular, open-source nature, SUMO is highly extensible.
An external control API (TraCI) enables users to observe and manipulate the simulation state via Python scripts, which opens up numerous opportunities, such as incorporating rebalancing algorithms.
While SUMO provides a high-fidelity representation of urban mobility suitable for evaluating performance under realistic conditions, it does not natively support agent-based modeling of users or vehicles, which must be implemented externally via TraCI.
\\

\subsubsection{ODySSEUS}
ODySSEUS is a simulation environment developed specifically for the operations of shared micromobility systems \cite{smartdatapolito_smartdatapolitoodysseus_2022}.
Its primary focus is modeling fleet and service dynamics, including demand generation, vehicle utilization, battery charging, and rebalancing operations.
While ODySSEUS does not simulate traffic interactions with the same level of detail as SUMO, it can incorporate simplified mobility models to approximate travel times.
Its main strength lies in modeling operational complexities, such as peak demand periods, staff schedules, and depot logistics for spare vehicles and battery swapping.
Losapio et al. \cite{losapio_smart_2022} employed an ODySSEUS-based environment to evaluate their ESB-DQN rebalancing framework, wrapping the simulator in a gym-compatible interface for training.
However, they noted that using ODySSEUS directly for training resulted in an overly long and expensive training process.
\\

\subsubsection{MATSim}
MATSim (Multi-Agent Transport Simulation) is an open-source, activity-based simulator that models travel demand and network loading at a macroscopic city scale.
It represents each traveler as an agent with a daily plan of activities, and simulates their mode choice and routes in an iterative manner.
Consequently, its level of detail is coarser than SUMO's, with traffic flow simulated using queue models rather than explicit vehicle tracking. MATSim has been utilized in several studies for evaluating rebalancing algorithms \cite{chouaki_review_2025}.
While its flexible architecture allows for custom modules, its Java-based implementation presents integration challenges for RL frameworks, which are predominantly developed in Python.
\\

\subsubsection{Summary}
Several platforms have been utilized in recent research to simulate shared micromobility fleets, facilitating the testing and evaluation of various rebalancing strategies.
SUMO is a microscopic traffic simulator that enables high-fidelity analysis of detailed vehicle interactions, energy consumption (especially relevant for electric fleets, considering factors such as elevation), and the localized traffic impacts of rebalancing operations.
However, implementing specific service logic and rebalancing strategies typically requires custom development via its TraCI interface.
ODySSEUS provides a specialized environment explicitly focused on the operational dynamics of shared fleets, leveraging real-world data to model and optimize fleet performance metrics, such as vehicle availability and rebalancing efficiency.
This makes it particularly suitable for data-driven analysis from a service provider's perspective.
Nevertheless, ODySSEUS simulates at a higher abstraction level than SUMO, which affects the realism of simulations performed using ODySSEUS.
In contrast, MATSim employs a large-scale, agent-based, mesoscopic approach centered on modeling traveler behavior and activity planning.
The program's strength lies in simulating how changes in service attributes affect user mode choice, overall travel patterns, demand satisfaction, and the system's balance.
This makes it ideal for strategic analysis and exploring user responses to incentives.
The implementation of custom rebalancing frameworks is more complex for MATSim since it is Java-based, whereas machine learning and similar algorithms are predominantly written in Python.

A notable trend in recent studies is the increasing use of custom simulation environments, particularly for RL approaches.
This is often necessary because integrating complex rebalancing algorithms into existing open-source simulators can be hindered by steep learning curves, insufficient documentation, and performance limitations, rendering them unsuitable for the iterative demands of RL training.
Especially, the increased computational overhead often makes it infeasible to implement such detailed simulations into the training process of rebalancing algorithms.
To mitigate the developmental and maintenance overhead of a separate high-fidelity environment, many studies consequently rely on a single, simplified simulator for both agent training and final validation.
Therefore, a notable research gap exists for a unified, high-fidelity simulation framework with a consistent interface.
Such a simulation environment could operate in a simplified mode for efficient training and a full-fidelity mode for accurate validation, thereby providing the benefits of realistic evaluation without the overhead of maintaining two distinct systems.
\\


\section{Summary and Research Gaps}
The preceding sections have reviewed the primary methodologies of micromobility fleet optimization through rebalancing algorithms.
These algorithms can be broadly categorized into mathematical optimization techniques, heuristics and meta-heuristics, and reinforcement learning.
Mathematical optimizations can provide optimal solutions to the rebalancing problem, but require a simplified static problem definition to be able to converge due to scalability issues.
In contrast, heuristics or meta-heuristics offer a computationally more efficient approach but can't guarantee optimality.
Reinforcement learning has emerged as a promising approach for handling the complex and dynamic nature of shared micromobility systems, although it introduces its individual challenges related to complexity.
Furthermore, the existing simulation tools for shared micromobility reveal a common trade-off between high-fidelity traffic simulation, specialized operational tools, and computational performance, leading to the most recent research developing custom simulators.
\\

From this comprehensive review, several distinct research gaps emerge. 
The most significant gap is the fragmented treatment of rebalancing modes.
Almost all examined research papers propose a system to optimize user-based or operator-based rebalancing in isolation, without a comprehensive framework that incorporates both rebalancing approaches.

Furthermore, a multitude of existing solutions, particularly those based on mathematical optimization or heuristics, follow a static nature.
They propose a rebalancing plan for an entire day, which is insufficient for shared micromobility systems, as demand experiences high fluctuations within a day, which such static rebalancing plans fail to account for.
Therefore, a clear requirement for rebalancing strategies is to develop dynamic solutions that can adjust to those demand fluctuations.

A third, more fundamental problem in this field is the use of historical data for driving the 
training and evaluation simulations.
This leads to the problem of censored demand.
Demand datasets only log completed trips; they fail to capture unmet demand, where a user cannot find an available vehicle within their acceptable walking radius.
Consequently, most models are optimized against an incomplete and biased demand model.
Addressing this data problem is a complex task, especially under the data availability conditions this thesis is facing (see \autoref{sec:design-data}).

In conclusion, this thesis addresses the gaps in integration and dynamism by proposing a novel framework that coordinates both user- and operator-based rebalancing on an hourly basis.
This approach aims to provide a more responsive and holistic solution to the fleet optimization problem, operating with the awareness that the underlying demand data represents a known simplification of reality.
\\