\chapter{Evaluation}\label{sec:evaluation} % approx. 8 pages
The following chapter details the experimental setup used for evaluation, the hyperparameter optimization for the RDC and UIC agents, and the 

\section{Experimental Setup}\label{sec:evaluation-setup}
Fleet sizes have been determined by the mean number of vehicles available in the bounds based on all GBFS snapshots.
For training and hyperparameter optimization, this fleet size has been reduced to 90\% to allow the agent to learn from more experiences with undersupply.
For the evaluation, three different fleet sizes were used: the mean from the GBFS data, 80\% of the mean, and 120\% of the mean.


\section{Hyperparameter Optimization}\label{sec:evaluation-ho}
The performance of reinforcement learning agents is highly dependent on the selection of their hyperparameters.
Therefore, a systematic hyperparameter optimization is essential for effective working agents, especially for a complex system like the proposed HRL framework.
This systematic hyperparameter optimization has been conducted for both the Regional Distribution Coordinator (RDC) agent and the User-Incentive Coordinator (UIC) agent to ensure high performance for each agent.
This section outlines the methodology employed for this optimization and presents the final parameter configurations used for evaluating the HRL framework.

The objective for the optimization of both agents is to maximize a composite reward function, which is a weighted sum of three metrics: maximizing user demand ($R_{demand}^{RDC}$; $R_{demand}^{UIC}$), minimizing rebalancing costs ($R_{cost}^{RDC}$; $R_{cost}^{UIC}$), and promoting equitable vehicle distribution ($R_{gini}^{RDC}$; $R_{gini}^{UIC}$).

For the HO process, the Optuna framework \cite{akiba_optuna_2019}, a state-of-the-art optimization tool, has been utilized.
The Tree-structured Parzen Estimator (TPE) sampler was selected as the search algorithm.
TPE is an optimization algorithm that models the search space probabilistically to efficiently find promising hyperparameter configurations.
The TPE sampler fits a Gaussian Mixture Model (GMM) to a set of parameters with the highest value for the objective function.
Then, a ratio between this GMM and another GMM fitted to the remaining parameters is calculated.
The sampler then chooses the parameter value that maximizes this ratio between the GMMs \cite{akiba_optuna_2019}.
This probabilistic approach makes TPE well-suited for complex and high-dimensional search spaces  \cite{akiba_optuna_2019}.

To ensure reproducibility of the HO process, all runs are seeded.
Since HO is computationally heavy, the HO process was parallelized across multiple jobs to accelerate the computation time.
Additionally, the demand forecasts were precalculated, which also decreased computation time.
The remaining section details the HO processes for the RDC and UIC agents.



\subsection{Regional Distribution Coordinator (RDC)}
The RDC agent, based on an MH-DQN architecture with PER, has 17 distinct hyperparameters.
To manage this complexity, a coarse-to-fine optimization strategy was adopted over three rounds.
For all rounds, each trial consists of 100 episodes, each with a maximum of 100 steps, and 20 evaluation episodes.\\

Initially, the parameters were grouped into five logical groups (Learning Rate, Architecture, Replay Buffer, Exploration Strategy, and Reward Weights).
The first round was conducted on a broad search space for each group independently across 50 trials.
This first round aimed to identify promising regions in the extensive search space.

In the second round, the search ranges for each parameter were refined based on the results of the previous round.
For parameters whose optimal parameters were found at the boundaries of the previous search space, the range was extended in this direction.
Conversely, for parameters whose optimal values were located centrally in the search space, the search space was narrowed down for more fine granular optimization.
This second, more focused round, was conducted over 50 trials per group as well.

In the final round, a unified fine-tuning sweep was performed across all 17 parameters simultaneously.
To combat the increase in complexity, the number of trials was increased to 100.
Additionally, the search spaces for each parameter were narrowed down based on the results from the previous round.
This final round ensures that dependencies between parameters of different groups are also addressed in the optimization process.\\

The resulting optimal hyperparameter configurations are presented in \autoref{tab:rdc_final_hyperparameters}.
More detailed descriptions for the parameters are provided in \autoref{tab:rdc_hyperparameters_description}, and the search spaces are defined in the \cref{tab:rdc_ho_first_round,tab:rdc_ho_second_round,tab:rdc_ho_final_round}.

\begin{table}[H]
    \centering
    \begin{tabularx}{\textwidth}{lXl}
        \toprule
        \textbf{Parameter} & \textbf{Summary Description} & \textbf{Final Value} \\
        \midrule
        \texttt{lr} & Initial learning rate & $1 \times 10^{-5}$ \\
        \texttt{lr\_step\_size} & LR decay step size & $1500$ \\
        \texttt{lr\_gamma} & LR decay factor & $0.9$ \\
        \texttt{gamma} & Discount factor & $0.926$\\
        \midrule
        \texttt{hidden\_size} & Hidden layer neuron count & $256$ \\
        \texttt{batch\_size} & Minibatch size & $256$ \\
        \texttt{tau} & Target network soft update rate & $0.001$ \\
        \midrule
        \texttt{rb\_capacity} & Replay buffer capacity & $10,000$\\
        \texttt{per\_alpha} & PER prioritization exponent ($\alpha$) & $0.8$\\
        \texttt{per\_beta\_start} & PER importance sampling start ($\beta_0$) & $0.3$\\
        \texttt{per\_beta\_frames} & PER $\beta$ annealing frames & $150,000$\\
        \midrule
        \texttt{epsilon\_start} & Initial exploration rate ($\epsilon_{start}$) & $1.0$\\
        \texttt{epsilon\_end} & Final exploration rate ($\epsilon_{end}$) & $0.06$\\
        \texttt{epsilon\_decay} & Exploration rate decay factor & $0.999$\\
        \midrule
        \texttt{weight\_demand} & Satisfied demand weight & $1.1$ \\
        \texttt{weight\_cost} & Rebalancing cost weight & $1.1$ \\
        \texttt{weight\_gini} & Gini index penalty weight & $0.1$ \\
        \bottomrule
    \end{tabularx}
    \caption{Final optimized hyperparameter configuration for the RDC.}
    \label{tab:rdc_final_hyperparameters}
\end{table}

\subsection{User-Incentive Coordinator (UIC)}
The User-Incentive Coordinator, which employs a PPO algorithm agent, has 15 different parameters for HO.
A multi-stage optimization strategy, similar to that used for the RDC agent, was implemented over three rounds.
Each trial of a round runs for 50,000 timesteps.\\

The initial round was conducted on a broad search space to identify search areas with high potential.
For the initial and second rounds, the hyperparameters were divided into four groups: PPO core, architecture, stability, and reward weights.

The second round used optimized search spaces for each parameter based on the results of the initial round.
The search spaces have been optimized similarly to the second round of the RDC's optimization strategy.

The third and last round optimized all parameters simultaneously to identify the most optimal hyperparameter configuration.
This final round ensures that the hyperparameters are optimized, even when accounting for dependencies between parameters from different groups.\\ 

The resulting optimal hyperparameter configurations for the UIC agent are presented in \autoref{tab:uic_final_hyperparameters}.
More detailed descriptions for the parameters are provided in \autoref{tab:uic_hyperparameters_description}, and the search spaces are defined in the \cref{tab:uic_ho_first_round,tab:uic_ho_second_round,tab:uic_ho_final_round}.

\begin{table}[!htb]
    \centering
    \begin{tabularx}{\textwidth}{lXl}
        \toprule
        Parameter & Description & Final Value \\
        \midrule
        \texttt{lr} & Optimizer learning rate. & $1 \times 10^{-6}$ \\
        \texttt{n\_steps} & Rollout buffer size per update. & $2048$ \\
        \texttt{batch\_size} & Minibatch size for gradient updates. & $32$ \\
        \texttt{clip\_range} & PPO policy update clipping range. & $0.13$\\
        \texttt{ent\_coeff} & Entropy bonus weight for exploration. & $0.00078$\\
        \texttt{vf\_coeff} & Value function loss weight. & $0.6$\\
        \midrule
        \texttt{gamma} & Reward discount factor. & $0.918$ \\
        \texttt{gae\_lambda} & Lambda parameter for GAE& $0.99$ \\
        \texttt{target\_kl} & Target KL divergence for policy updates. & $0.02$\\
        \midrule
        \texttt{num\_hidden\_layers} & Number of network hidden layers. & $3$\\
        \texttt{hidden\_size} & Neurons per hidden layer. & $256$ \\
        \texttt{activation\_fn} & Hidden layer activation function. & Tanh \\
        \midrule
        \texttt{weight\_demand} & Satisfied demand weight & $1.1$ \\
        \texttt{weight\_cost} & Rebalancing cost weight & $1.1$ \\
        \texttt{weight\_gini} & Gini index penalty weight & $0.1$ \\
        \bottomrule
    \end{tabularx}
    \caption{Final hyperparameter values for the User-Incentive Coordinator after optimization.}
    \label{tab:uic_final_hyperparameters}
\end{table}

\section{Results and Analysis}